Supermassive Black Hole Dynamics: A Proposal for a New Class of X-Ray Sources

### 2019-04-05 paper.tex
More Jerry edits; changed title of paper

\title{Cannibalism among the supermassive black holes}

@@

\begin{abstract}
We perform N-body simulations on 13 galaxies extracted from a cosmological simulation of hierarchical structure formation with total mass in the range $10^{12} \msun < M < 10^{13} \msun$ from $4 \ge z \ge 0$.  We track the dynamical evolution of the orbiting black holes (BHs) around their host central BH.  Of the 120 orbiting BHs, we find that 50 merge with their host central BHs, 11 are ejected from their respective host galaxies, 59 are still orbiting at the end of the simulations.  Across all galaxies, 48 BHs are kicked to a higher orbit after close interaction with the central BH, after which none of them merge with their hosts.  We calculate the x-ray luminosities of the orbiting massive BHs at z=0, and find a range of $10^{18}-10^{37}$ $erg\ s^{-1}$.
\end{abstract}

@@

How common are mergers of massive black holes?  For the galaxies themselves we know that minor mergers are frequent from the observed (\citet{2010ApJ...718L..73V}, \citet{2008ApJ...677L...5V}, \citet{2019MNRAS.484..595M}) evolution of the size and mass of these systems, but assuming that all galaxy mergers lead to SMBH mergers overpredicts the observations of gravitational waves from the pulsar timing arrays (PTAs) (\citet{2008MNRAS.390..192S}, \citet{2009MNRAS.394.2255S}, \citet{2013MNRAS.433L...1S}, \citet{2014ApJ...789..156M}, \citet{2015ApJ...799..178K}, \citet{2018ApJ...856...42S}, \citet{2018arXiv180403143I}).  However, we do not often see multiple SMBHs at the centers of massive ellipticals.  So, what does happen?  One paper (\citet{2018MNRAS.473.3410R}) has indicated that dynamical interactions among multiple black holes, which eject a non-negligible fraction of the mass, may solve this problem.  The present paper also addresses this purported dynamical solution, focusing attention on the large fraction of lower mass black holes that remain to be detected as they orbit in massive galaxies.  What we argue in this paper is that, as a result of the N-body interactions among the infalling black holes, some are ejected, some remain in extended orbits and some (the few most massive ones) do in fact merge, but these do not exceed the PTA limits, and the orbiting ones should be detectable via their ability to accrete gas and emit radiation.

@@

This paper is organized as follows: in Section \ref{sec:methods} we describe the cosmological simulations from which we use the merger history to set up our idealized numerical simulations. We present the few-body integration code, \textsc{AR-Chain} that we used for our simulations of SMBH dynamics, and the modifications we made to this code in order to deal with a host galaxy's gravitational potential. In Section \ref{sec:results}, we show the results of our XX exemplary simulations of galaxies growing with time and acquiring new SMBHs. We analyze how the SMBHs are driven into the core of their new host galaxies and how interaction with the host black hole leads to near-ejections or mergers. We then estimate the X-ray luminosities expected of the orbiting black holes.  The final Section \ref{sec:conclusions} contains a discussion of the results and our conclusions.

@@

We can identify Equation \ref{eq:df} as the dynamical friction term. The second term introduces a variance of the friction term, and even allows the SBHs to be accelerated when the velocity of a SBH is sufficiently small. The third term introduces a change in velocity perpendicular to the direction of motion of the SBH. It is a randomly oriented vector, and hence causes the SBHs to execute a random walk in phase space. The last two terms will establish that the SBHs are ultimately in energy equipartition with the background stars.

@@

The change of energy, $\mbox{d}E_{BH}$, of the orbiting black hole due to phase-space diffusion is given back to the stellar background potential, with $\mbox{d}E = -\mbox{d}E_{BH}$. As a consequence of this energy transfer, inspiralling black holes will cause an expansion of the NSC. For this purpose we calculate the change in potential energy, $\mbox{d}W$, of the stellar system using

@@

\subsubsection{Mergers}\label{sec:mergers}
The tendency of larger mass black holes to merge with their host can perhaps be more clearly seen by looking at the merger mass ratio between the incoming black hole and the host.  Figure \ref{fig:q_merge_fraction} displays the fraction of galaxy mergers that result in black hole mergers versus the merger mass ratio, defined as $m_2/m_1$, where $m_2$ is the orbiting black hole and $m_1$ is the host black hole.  The plot monotonically increases with increasing mass ratio, reaching nearly unity for mass ratio of $10^{-1}$, meaning nearly all galaxy mergers with a black hole mass ratio of $10^{-1}$ resulted in the orbiting black hole merging with the host.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/"Final Radius v Mass of Orbiting BHs_corrected_".png}
\caption{Final Radii of remaining orbital black holes at z=0}
\label{fig:mvr}
\end{center}
\end{figure}

>>>>

\title{Supermassive Black Hole Dynamics: A Proposal for a New Class of X-Ray Sources}

@@

\begin{abstract}
We perform N-body simulations on 13 galaxies extracted from a cosmological simulation of hierarchical structure formation with total mass in the range $10^{12} \msun < M < 10^{13} \msun$ from $4 \ge z \ge 0$.  We track the dynamical evolution of the orbiting black holes (BHs) around their host central BHs.  Of the 120 orbiting BHs, we find that 50 merge with their host central BHs, 11 are ejected from their respective host galaxies, 59 are still orbiting at the end of the simulations.  Across all galaxies, 48 BHs are kicked to a higher orbit after close interactions with the central BH, after which none of them merge with their hosts.  These orbiting BHs should be detectable by their anomalous (not LMXRB) spectra.  The x-ray luminosities of the orbiting massive BHs at z=0 are in the range $10^{23}-10^{37}$ $erg\ s^{-1}$, with median value $10^{29}$ $erg\ s^{-1}$.
\end{abstract}

@@

How common are mergers of massive black holes?  For the galaxies themselves we know that minor mergers are frequent from the observed (\citet{2010ApJ...718L..73V}, \citet{2008ApJ...677L...5V}, \citet{2019MNRAS.484..595M}) evolution of the size and mass of these systems, but assuming that all galaxy mergers lead to SMBH mergers overpredicts the observations of gravitational waves from the pulsar timing arrays (PTAs) (\citet{2008MNRAS.390..192S}, \citet{2009MNRAS.394.2255S}, \citet{2013MNRAS.433L...1S}, \citet{2014ApJ...789..156M}, \citet{2015ApJ...799..178K}, \citet{2018ApJ...856...42S}, \citet{2018arXiv180403143I}).  However, we do not often see multiple SMBHs at the centers of massive ellipticals.  So, what does happen?  One paper (\citet{2018MNRAS.473.3410R}) has indicated that dynamical interactions among multiple orbiting black holes, which will eject a non-negligible fraction of the mass, may solve this problem.  The present paper also addresses this purported dynamical solution, focusing attention on the large fraction of lower mass black holes that remain to be detected as they orbit in massive galaxies.  What we argue in this paper is that, as a result of the N-body interactions among the infalling black holes, some are ejected, some remain in extended orbits and some (the few most massive ones) do in fact merge, but these do not exceed the PTA limits, and the orbiting ones should be detectable via their ability to accrete gas and emit radiation.

@@

This paper is organized as follows: in Section \ref{sec:methods} we describe the cosmological simulations from which we use the merger history to set up our idealized numerical simulations. We present the few-body integration code, \textsc{AR-Chain} that we used for our simulations of SMBH dynamics, and the modifications we made to this code in order to deal with a host galaxy's gravitational potential. In Section \ref{sec:results}, we show the results of our XX exemplary simulations of galaxies growing with time and acquiring new SMBHs. We analyze how the SMBHs spiral into the core of their new host galaxies due to dynamical friction, and how interactions with the host black hole and other orbiting BHs leads to near-ejections or mergers. We then estimate the X-ray luminosities expected of the orbiting black holes.  The final Section \ref{sec:conclusions} contains a discussion of the results and our conclusions.

@@

We can identify Equation \ref{eq:df} as the dynamical friction term. The second term introduces a variance of the friction term, and even allows the SBHs to be accelerated when the velocity of a SBH is sufficiently small. The third term introduces a change in velocity perpendicular to the direction of motion of the SBH. It is a randomly oriented vector, and hence causes the SBHs to depart from smooth orbits due to the small random velocity kicks. The last two terms will establish that the SBHs are ultimately in energy equipartition with the background stars.

@@

The change of energy, $\mbox{d}E_{BH}$, of the orbiting black hole due to phase-space diffusion is given back to the stellar background potential, with $\mbox{d}E = -\mbox{d}E_{BH}$. As a consequence of this energy transfer, inspiralling black holes will cause an expansion of the stellar system. For this purpose we calculate the change in potential energy, $\mbox{d}W$, of the stellar system using

@@

\subsubsection{Mergers}\label{sec:mergers}
The tendency of larger mass black holes to merge with their host can perhaps be more clearly seen by looking at the merger mass ratio between the incoming black hole and the host.  Figure \ref{fig:q_merge_fraction} displays the fraction of galaxy mergers that result in black hole mergers versus the merger mass ratio, defined as $m_2/m_1$, where $m_2$ is the orbiting black hole and $m_1$ is the host black hole.  The plot monotonically increases with increasing mass ratio, reaching nearly unity for mass ratio of $10^{-1}$, meaning nearly all galaxy mergers with a black hole mass ratio greater than $10^{-1}$ resulted in the orbiting black hole merging with the host.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.45\textwidth]{plots/"Final Radius v Mass of Orbiting BHs".png}
\caption{Final Radii of remaining orbital black holes at z=0}
\label{fig:mvr}
\end{center}
\end{figure}
